Exploring the Validity Range of PLL Analytical Models

### 2025-05-29 CaseStudies/Example3_2_LimitsoftheTheorySoFar/Report/article/article.qmd
Reformulating the article.

# Analysis

For the analysis of the linear transient behavior of a PLL, the phase error, rather than the output phase, is needed. So, from the [Fig. @fig-PLL-integerN-blocks], it's possible to get the relation between the phase error, $\theta_e$, and the input phase reference, $\theta_R$. This transfer function can be expressed as:

$$
\begin{split}
\frac{\theta_e}{\theta_R}&=\frac{s^2}{s^2+2\zeta\omega_ns+\omega_n^2}
\end{split}
$$ {#eq-phase-error-transfer-function}

@@

The example whose the resolution is being given here is studying the effect of an input frequency step. Since the input is described by phase, as it's possible to be seen in [Eq. @eq-phase-error-transfer-function] ($\theta_R$), it's possible to take the phase equivalent of a frequency step, which is a phase ramp, since the phase is the integral of frequency, and the integral of a step is a ramp. Thus, the input can be described as:

@@

$$
\begin{split}
\theta_e&=\frac{\Delta\omega}{s^2+2\zeta\omega_ns+\omega_n^2}
\end{split}
$$

And getting the inverse Laplace transform, for $\zeta<1$, results in:

@@

@@

@@

Thus, the natural frequency is $\omega_n=$ `{python} format_with_pi_latex(omega_n)`rad/s.

## Maximum Phase Error Instant

@@

## Maximum Normalized Phase Error

@@

## Maximum Allowed Frequency Step

The goal is to find the maximum input frequency step, $\Delta\omega_m$, such that the maximum phase error does not exceed $\theta_{e_m}=$ `{python} format_with_pi_latex(theta_e_m)`. Using the normalized result:

$$
\begin{split}
\theta_{e_m}&=\frac{\Delta\omega_m}{\omega_n}\theta_{e_\text{m,n}}\implies\\
\Delta\omega_m&=\frac{\theta_{e_m}\omega_n}{\theta_{e_\text{m,n}}}
\end{split}
$$

@@

Substituting the calculated values yields:

- $\Delta\omega_m=$ `{python} format_with_pi_latex(step_m_ang)`rad/s
- $\Delta f_m=$ `{python} format(step_m_lin)`Hz

Hence, the maximum input frequency step that still allows theoretical predictability is `{python} format(step_m_lin)`Hz. If the input frequency step exceeds this limit, the PLL may lose lock and exhibit cycle slipping (more details are provided in the next section).

## Settling Time Estimation

@@

This results in a settling time of $t_s=$ `{python} format(t_s)`s.

# Cycle Slipping

# Simulation

@@

@@

omega_n = omega_bw / np.sqrt(1 + 2 * zeta**2 + np.sqrt(4 * zeta**4 + 4 * zeta**2 + 2))
f_n = omega_n/2/np.pi

@@

@@

theta_e_m_meas = 2 * math.pi

@@

F_sys = signal.TransferFunction(F_num, F_den)

# F(s) * K / s = G(s)
G_num = np.polymul(F_num, [K])
G_den = np.polymul(F_den, [1, 0])  # Multiplica por s

# H(s) = G(s) / (1 + G(s))
H_num = G_num
H_den = np.polyadd(G_den, G_num)

# Criando o sistema de tempo contínuo
H_sys = signal.TransferFunction(H_num, H_den)

@@

@@

# Criar a função de transferência final
E_sys = signal.TransferFunction(num_E, den_E)

@@

@@

# Escalar do denominador: (1/N) * Kvco
den_scalar = (1 / N) * Kvco

# Para dividir por escalar, basta dividir o numerador
num_Vc = [coef / den_scalar for coef in H_sys.num]
den_Vc = H_sys.den

# Criar nova função de transferência
Vc_sys = signal.TransferFunction(num_Vc, den_Vc)

@@

Using the parameters $\omega_n$ and $\zeta$, the transfer function of the entire loop can be described as (neglecting the second capacitor, $C_2$, of the filter):

$$
\frac{\theta_o}{\theta_R}=\frac{\omega_n^2\left(\frac{2\zeta}{\omega_n}s+1\right)}{s^2+2\zeta\omega_n s+\omega_n^2}
$$ {#eq-full-transfer-function-no-C2}

An integer-N Phase-Locked Loop (PLL) was developed with a division factor of $N=$ `{python} N`. The system was designed with a VCO gain of $K_\mathrm{VCO}=$ `{python} format(KVCO)`Hz/V, and a charge pump current of $I=$ `{python} format(I)`A. The phase detector gain is given by $K_\theta=I/2\pi$, resulting in $K_\theta=$ `{python} format(Kphase)`A/rad.

Based on the chosen parameters, the loop filter components were calculated as follows. The primary capacitor of the loop filter is:

$$
C_1 = \frac{K_\theta K_\mathrm{VCO}}{N \omega_n^2}
$$

That is, $C_1=$ `{python} format(C1)`F.

The second capacitor is chosen to be one-tenth of $C_1$, i.e., $C_2=$ `{python} format(C2)`F.

The resistor of the loop filter is given by:

$$
R_1 = \frac{\zeta 4 \pi N \omega_n}{I K_\mathrm{VCO}}
$$

That is, $R_1=$ `{python} format(R1)`$\Omega$.

Similarlly what was presented in [Eq. @eq-full-transfer-function-no-C2], it's possible to present the transfer function of the system in terms of components values. This transfer function can be seen below (this equation uses the second capacitor, $C_2$, of the loop filter, adding a high-frequency pole to the system):

where $C_s$ is the paralel between $C_1$ and $C_2$. With the PLL values properly defined, the transfer function of this system is given by:

def poly_to_str(coeffs):
    terms = []
    degree = len(coeffs) - 1
    for i, coef in enumerate(coeffs):
        power = degree - i
        if abs(coef) < 1e-12:  # ignora coeficientes muito pequenos
            continue
        term = ""
        coef_str = format_eng(coef)
        if not (abs(coef - 1) < 1e-12 and power != 0):
            term += coef_str
        if power > 0:
            term += "s"
            if power > 1:
                term += f"^{power}"
        terms.append(term)
    return " + ".join(terms)

# Em scipy.signal, num e den são arrays dentro de listas (mesmo para SISO)
num = H_sys.num[0][0] if isinstance(H_sys.num[0], (list, tuple)) else H_sys.num
den = H_sys.den[0][0] if isinstance(H_sys.den[0], (list, tuple)) else H_sys.den

# Exibe como equação no Jupyter
display(Markdown(f"$$H(s) = \\frac{{{num_str}}}{{{den_str}}}$$ {{#eq-transfer-function}}"))

def poly_to_str(coeffs):
    terms = []
    degree = len(coeffs) - 1
    for i, coef in enumerate(coeffs):
        power = degree - i
        if abs(coef) < 1e-12:  # ignora coeficientes muito pequenos
            continue
        term = ""
        coef_str = format_eng(coef)
        if not (abs(coef - 1) < 1e-12 and power != 0):
            term += coef_str
        if power > 0:
            term += "s"
            if power > 1:
                term += f"^{power}"
        terms.append(term)
    return " + ".join(terms)

# Em scipy.signal, num e den são arrays dentro de listas (mesmo para SISO)
num = F_sys.num[0][0] if isinstance(F_sys.num[0], (list, tuple)) else F_sys.num
den = F_sys.den[0][0] if isinstance(F_sys.den[0], (list, tuple)) else F_sys.den

# Exibe como equação no Jupyter
display(Markdown(f"$$F(s) = \\frac{{{num_str}}}{{{den_str}}}$$ {{#eq-transfer-function-loop-filter}}"))

@@

Using the Simulink software, a testbench was prepared to verify the settling time of the PLL. With the previously calculated parameters and using the maximum step frequency value, the resulting curve can be seen in [Fig. @fig-max-step-response]. It can be observed that the step is applied at the instant of 1 ms, as the system is already properly stabilized at that moment. As predicted, cycle slipping does not occur; however, unlike what was calculated, the settling time of this signal is $9.196\;\mu\text{s}$.

![System response to the maximum step frequency](../images/max_step_frequency_response.jpg){#fig-max-step-response}

```{python}
#| echo: false
cycle_slip_step_lin = 3*step_m_lin

Moreover, contrary to what was previously predicted, for cycle slipping to occur, the step frequency must be approximately $\Delta f=$ `{python} format(cycle_slip_step_lin)`Hz. The system's response to this signal can be seen in [Fig. @fig-cycle-slip-response].

![System response with cycle slip](../images/step_frequency_response_cycle_slip.jpg){#fig-cycle-slip-response}

>>>>

# System Modeling

A [Fig. @fig-PLL-integerN-blocks] o diagrama de blocos de um PLL Integer-N PFD/CP-based.

@@

The loop filter transfer function, $F(s)$, neglecting the shunt capacitor, $C_2$ can be described as:

$$
F(s) = R+\frac{1}{sC_1}=\frac{sC_1R+1}{sC_1}
$$

Ainda ignorando o segundo capacitor, é possível escrever a função de transferência de malha aberta desse sistema apresentado na [Fig. @fig-PLL-integerN-blocks] como

$$
G(s) = \frac{KF(s)}{s}
$$

em que

$$
K = A_0K_\theta K_\mathrm{VCO}N^{-1}
$$

tal que $A_0$ é o ganho DC, $K_\theta=I/2\pi$ o ganho do charge pump, $K_\mathrm{VCO}$ o ganho do VCO e $N$ o divisor de frequência.

A função de transferência em malha fechada é dada por

$$
\begin{split}
H(s)&=\frac{\theta_o}{\theta_R}=\frac{G(s)}{1+G(s)}\implies\\
&=\frac{\frac{A_0K_\theta K_\mathrm{VCO}}{N}\left(R+\frac{1}{sC_1}\right)}{s+\frac{A_0K_\theta K_\mathrm{VCO}}{N}\left(R+\frac{1}{sC_1}\right)}
\end{split}
$$ {#eq-PLL-closed-loop-transfer-function-initial}

That can be rearanged as:

$$
H(s)=\frac{\frac{A_0K_\theta K_\mathrm{VCO}}{2\pi\cdot NC_1}\left(RC_1s+1\right)}{s^2+\frac{A_0K_\theta IK_\mathrm{VCO}}{2\pi\cdot N}Rs+\frac{A_0K_\theta IK_\mathrm{VCO}}{2\pi\cdot NC_1}}
$$ {#eq-PLL-closed-loop-transfer-function-modified}

ou, de maneira mais simples,

$$
H(s)=\frac{KF(s)}{s+KF(s)}
$$

Com o filtro simplificado, sem o secundo capacitor,

$$
H(s)=\frac{\frac{K}{C_1}\left(RC_1s+1\right)}{s^2+KRs+\frac{K}{C_1}}
$$

Using the parameters $\omega_n$ and $\zeta$, the closed-loop transfer function of the entire loop can be described as (neglecting the second capacitor, $C_2$, of the filter which would yield to a not wished third-order transfer function):

$$
H(s)=\frac{\omega_n^2\left(\frac{2\zeta}{\omega_n}s+1\right)}{s^2+2\zeta\omega_n s+\omega_n^2}
$$ {#eq-PLL-closed-loop-transfer-function-zeta-omega_n}

with

$$
\omega_n = \sqrt{\frac{K}{C_1}}
$$ {#eq-omega_n-definition-no-C2}

and

$$
\zeta = \frac{R}{2}\sqrt{KC_1}
$$ {#eq-zeta-definition-no-C2}

Thus the components of the loop filter can be obtained as:

$$
C_1 = \frac{K}{\omega_n^2}
$$ {#eq-C1-loop-filter}

and

$$
R_1 = \zeta\sqrt{\frac{2\omega_n}{K}}
$$ {#eq-R1-loop-filter}

For simplicity, the second capacitor can be obtained by $C_2=C_1/10$. Se esse segundo capacitor não fosse desconsiderado, os passos aqui apresentados seriam um pouco mais complicados. Entretanto, ainda it's possible to present the transfer function of the system in terms of all components values. This transfer function can be seen below (this equation uses the second capacitor, $C_2$, of the loop filter, adding a high-frequency pole to the system):

where $C_s$ is the paralel between $C_1$ and $C_2$.

## Phase Error Transfer Function

A função de transferência do erro de fase pela fase de referência pode ser obtida como

$$
E(s)=\frac{\theta_e}{\theta_R}=H(s)\cdot\frac{s}{KF(s)}
$$ {#eq-phase-error-transfer-function-initial}

Utilizando as expressões [Eq. @eq-PLL-closed-loop-transfer-function-zeta-omega_n] e [Eq. @eq-PLL-closed-loop-transfer-function-modified],

$$
\begin{split}
E(s)&=\frac{\frac{A_0K_\theta K_\mathrm{VCO}}{NC_1}\left(RC_1s+1\right)}{s^2+2\zeta\omega_ns+\omega_n^2}\cdot\frac{1}{1/N}\cdot\frac{1}{A_0K_\mathrm{VCO}/s}\cdot\frac{1}{F(s)}\cdot\frac{1}{K_\theta}\implies\\
&=\frac{\frac{A_0K_\theta K_\mathrm{VCO}}{NC_1}\left(RC_1s+1\right)}{s^2+2\zeta\omega_ns+\omega_n^2}\cdot N\cdot\frac{s}{A_0K_\mathrm{VCO}}\cdot\frac{sC_1}{RC_1s+1}\cdot\frac{1}{K_\theta}
\end{split}
$$

Resulting in:

$$
E(s)=\frac{s^2}{s^2+2\zeta\omega_ns+\omega_n^2}
$$ {#eq-phase-error-transfer-function}

The example whose the resolution is being given here is studying the effect of an input frequency step. Since the input is described by phase, as it's possible to be seen in [Eq. @eq-phase-error-transfer-function-initial] ($\theta_R$), it's possible to take the phase equivalent of a frequency step, which is a phase ramp, since the phase is the integral of frequency, and the integral of a step is a ramp. Thus, the input can be described as:

@@

$$
\begin{split}
\theta_e&=\frac{\Delta\omega}{s^2+2\zeta\omega_ns+\omega_n^2}
\end{split}
$$ {#eq-transfer-function-step-frequency-phase-error}

Getting the inverse Laplace transform of [Eq. @eq-transfer-function-step-frequency-phase-error], for $\zeta<1$, results in:

@@

### Maximum Phase Error Instant

### Maximum Normalized Phase Error

### Maximum Allowed Frequency Step

The goal is to find the maximum input frequency step, $\Delta\omega_m$, such that the phase error does not exceed the maximum phase error that the PFD can withstand. Usually, this value is about $2\pi$. Using the normalized result:

$$
\theta_{e_m}=\frac{\Delta\omega_m}{\omega_n}\theta_{e_\text{m,n}}
$$

Levando a um máximo step frequency igual a

$$
\Delta\omega_m=\frac{\theta_{e_m}\omega_n}{\theta_{e_\text{m,n}}}
$$ {#eq-maximum-step-frequency}

### Settling Time Estimation

## Voltage Control Transfer Function

A função de transferência da tensão de controle pela fase de referência pode ser obtido como

$$
V(s)=\frac{V_C}{\theta_R}=H(s)\cdot\frac{1}{1/N}\cdot\frac{1}{K_\mathrm{VCO}/s}
$$ {#eq-voltage-control-transfer-function-initial}

Utilizando as expressões [Eq. @eq-PLL-closed-loop-transfer-function-zeta-omega_n] e [Eq. @eq-PLL-closed-loop-transfer-function-modified],

$$
\begin{split}
V(s)&=\frac{\frac{A_0K_\theta K_\mathrm{VCO}}{NC_1}\left(RC_1s+1\right)}{s^2+2\zeta\omega_ns+\omega_n^2}\cdot\frac{1}{1/N}\cdot\frac{1}{K_\mathrm{VCO}/s}\implies\\
&=\frac{\frac{A_0K_\theta K_\mathrm{VCO}}{NC_1}\left(RC_1s+1\right)}{s^2+2\zeta\omega_ns+\omega_n^2}\cdot N\cdot\frac{s}{K_\mathrm{VCO}}
\end{split}
$$

Resulting in:

$$
V(s)=\frac{\frac{A_0K_\theta}{C_1}s\left(RC_1s+1\right)}{s^2+2\zeta\omega_ns+\omega_n^2}
$$ {#eq-voltage-control-transfer-function}

Multiplicando o step de frequência nessa função, é obtido

$$
\begin{split}
V_C(s)&=\frac{\Delta\omega\frac{A_0K_\theta}{C_1s}\left(RC_1s+1\right)}{s^2+2\zeta\omega_ns+\omega_n^2}
\end{split}
$$ {#eq-transfer-function-step-frequency-voltage-control}

@@

## Cycle Slipping

# Design Analysis

@@

Partindo dos parâmetros já apresentados na introdução, the natural frequency is $\omega_n=$ `{python} format_with_pi_latex(omega_n)`rad/s.

@@

Calculando-se o instante em que ocorre a máxima diferença de fase obtém-se, $t_m=$ `{python} format(t_m)`s.

@@

Normalmente, espera-se que o máximo erro de fase suportado pelo PFD é de $2\pi$, entretanto, para coincidir com os valores obtidos com o simulador utilizado, empiricamente obtém-se um maximum phase error, $\theta_{e_m}=$ `{python} format_with_pi_latex(theta_e_m)`.

@@

Substituting the calculated values na [Eq. @eq-maximum-step-frequency] yields $\Delta\omega_m=$ `{python} format_with_pi_latex(step_m_ang)`rad/s or, alternativelly, $\Delta f_m=$ `{python} format(step_m_lin)`Hz. Hence, the maximum input frequency step that still allows theoretical predictability is `{python} format(step_m_lin)`Hz. If the input frequency step exceeds this limit, the PLL may lose lock and exhibit cycle slipping (more details are provided in the next section).

@@

Com esse step de frequencia e todos os parâmetros calculados até então, é possível calcular o settling time de 1% de solução, resultando em $t_s=$ `{python} format(t_s)`s.

# System Responses

@@

An integer-N Phase-Locked Loop (PLL) was developed with a division factor of $N=$ `{python} N`. The system was designed with a VCO gain of $K_\mathrm{VCO}=$ `{python} format(KVCO)`Hz/V, and a charge pump current of $I=$ `{python} format(I)`A. The phase detector gain is given by $K_\theta=I/2\pi$. Além disso, para que seja possível coincidir os valores calculados com as respostas obtidas na simulação, foi obtido um valor empírico para o ganho dc, $A_0=$ `{python} round(A0, 2)`.

@@

omega_n = omega_bw / np.sqrt(1 + 2 * zeta**2 + np.sqrt(4 * zeta**4 + 4 * zeta**2 + 2))
f_n = omega_n/2/np.pi
```

@@

Based on the chosen parameters, the loop filter components were calculated as follows. The primary capacitor of the loop filter is given by the [Eq. @], that is, $C_1=$ `{python} format(C1)`F. The second capacitor is chosen to be one-tenth of $C_1$, i.e., $C_2=$ `{python} format(C2)`F. The resistor of the loop filter is given by [Eq. @], that is, $R_1=$ `{python} format(R1)`$\Omega$.

@@

theta_e_m_meas = theta_e_m

@@

F_sys = signal.TransferFunction(F_num, F_den)
```

With the PLL values properly defined, the transfer function of this system is given by:

def poly_to_str(coeffs):
    terms = []
    degree = len(coeffs) - 1
    for i, coef in enumerate(coeffs):
        power = degree - i
        if abs(coef) < 1e-12:  # ignora coeficientes muito pequenos
            continue
        term = ""
        coef_str = format_eng(coef)
        if not (abs(coef - 1) < 1e-12 and power != 0):
            term += coef_str
        if power > 0:
            term += "s"
            if power > 1:
                term += f"^{power}"
        terms.append(term)
    return " + ".join(terms)
```

```{python}
#| echo: false
# Em scipy.signal, num e den são arrays dentro de listas (mesmo para SISO)
num = F_sys.num[0][0] if isinstance(F_sys.num[0], (list, tuple)) else F_sys.num
den = F_sys.den[0][0] if isinstance(F_sys.den[0], (list, tuple)) else F_sys.den

# Exibe como equação no Jupyter
display(Markdown(f"$$F(s) = \\frac{{{num_str}}}{{{den_str}}}$$ {{#eq-filter-loop-transfer-function}}"))

@@

```{python}
#| echo: false
# F(s) * K / s = G(s)
G_num = np.polymul(F_num, [K])
G_den = np.polymul(F_den, [1, 0])  # Multiplica por s

G_sys = signal.TransferFunction(G_num, G_den)
```

With the PLL values properly defined, the transfer function of this system is given by:

```{python}
#| echo: false
# Em scipy.signal, num e den são arrays dentro de listas (mesmo para SISO)
num = G_sys.num[0][0] if isinstance(G_sys.num[0], (list, tuple)) else G_sys.num
den = G_sys.den[0][0] if isinstance(G_sys.den[0], (list, tuple)) else G_sys.den

# Exibe como equação no Jupyter
display(Markdown(f"$$G(s) = \\frac{{{num_str}}}{{{den_str}}}$$ {{#eq-open-loop-num-transfer-function}}"))
```

```{python}
#| echo: false
# H(s) = G(s) / (1 + G(s))
H_num = G_num
H_den = np.polyadd(G_den, G_num)

# Criando o sistema de tempo contínuo
H_sys = signal.TransferFunction(H_num, H_den)
```

With the PLL values properly defined, the closed-loop transfer function of this system is given by:

```{python}
#| echo: false
# Em scipy.signal, num e den são arrays dentro de listas (mesmo para SISO)
num = H_sys.num[0][0] if isinstance(H_sys.num[0], (list, tuple)) else H_sys.num
den = H_sys.den[0][0] if isinstance(H_sys.den[0], (list, tuple)) else H_sys.den

# Exibe como equação no Jupyter
display(Markdown(f"$$H(s) = \\frac{{{num_str}}}{{{den_str}}}$$ {{#eq-closed-loop-num-transfer-function}}"))
```

@@

# Criar a função de transferência final
E_sys = signal.TransferFunction(num_E, den_E)
```

@@

# Vc_sys = H_sys / ((1/N)*(Kvco/s))
#Kvco_over_s = ([Kvco], [1, 0])
Kvco_over_s = ([Kvco], [1])
den_num = np.convolve([1/N], Kvco_over_s[0])
den_den = Kvco_over_s[1]

num_Vc = np.convolve(H_sys.num, den_den)
den_Vc = np.convolve(H_sys.den, den_num)

Vc_sys = signal.TransferFunction(num_Vc, den_Vc)
```

@@

@@

@@

# Behavioral Simulation
